Hierarchical recurrent models reveal latent states of neural activity in \textit{C. elegans}

### 2017-12-21 drafts/elife/draft.tex
emission model results

\begin{figure}[ht]
\centering%
\includegraphics[width=6in]{figures/figure1} 
\caption{A schematic of the hierarchical model for combining
  information across multiple worms in order to learn a shared
  dynamical system for \celegans.  \textbf{a.}~All worms share a
  canonical dynamics library. Here, we have three dynamics functions
  (blue, red, gold), each of which is a simple linear mapping from
  $x_t \in \reals^2$ to $x_{t+1} \in \reals^2$.  The recurrent model
  assigns probabilities to the different dynamics functions based on
  the current location in $x$-space.  \textbf{b.}~The worms also share
  an emission matrix, $C$, which can be viewed as an embedding of
  neurons in latent space.  \textbf{c.}~Each worm has its own dynamics
  library, which allows for slight deviations from the canonical
  dynamics.  \textbf{d.}~The model segments neural activity into a
  sequence of discrete states~$z_{t} \in \{1, 2, 3\}$, which index
  into the dynamics library.  \textbf{e.}~The discrete states specify
  the dynamics that are used to propagate the continuous latent
  states~$x_t$ forward in time. Here, the two dimensional latent
  states are colored according to the instantaneous discrete
  state. \textbf{f.}~The observed neural activity is a linear
  projection of the underlying continuous latent states.  On any given
  trial, we observe only a fraction of all the neurons.  The
  hierarchical model combines these partial recordings to infer the
  underlying states, the dynamics libraries, and the shared neural
  embedding. }
\label{fig:model}
\end{figure}

@@

@@

% Robust model
While recurrent SLDS can approximate highly nonlinear systems, they
are still misspecified. \emph{Robust} methods~\citep{huber1981robust}
ameliorate this concern by lessening the penalty on observations that
do not quite match the dynamics of the current regime.  We achieve
this robustness by replacing the standard Gaussian noise model with a
multivariate-t distribution~\citep{lange1989robust}.  The heavy tails
of this distribution permit outliers and account for some of the model
misspecification.

% Hierarchical model
The third component of our framework is a \emph{hierarchical} model
for learning canonical dynamics from partial recordings collected from
many worms.  Since each recording contains a different subset of
labeled neurons, our first step is to align the data, treating the
activity of neurons that were not found in a given recording as
missing data that must be inferred.  We handle this by integrating
over possible activity of the missing neurons, analytically
marginalizing out this missing data.  The second challenge is
combining information from many recordings while also allowing for
worm-to-worm differences.  For example, the neural activitiy may have
larger variance in some recordings than in others due to different
levels of expression of the fluorescent protein.  We allow for this by
having a separate variance for each worm and neuron, as shown in
Fig.~\ref{fig:model}f.  More importantly, the dynamics of neural
activity may be slightly different from worm to worm.  We account for
this by introducing a canonical library of dynamical regimes
(Fig.~\ref{fig:model}e) that are shared by all worms as well as a
unique library for each individual worm (Fig.~\ref{fig:model}d);
however, we constrain the per-worm dynamics to be close to the
canonical dynamics.

\subsection*{Emission models reveal clusters of neurons and aid in identification}

\begin{figure}[t]
\centering%
\includegraphics[width=6in]{figures/figure2} 
\caption{}
\label{fig:emissions}
\end{figure}

>>>>

@@

\begin{figure}[t!]
\centering%
\includegraphics[width=6in]{figures/figure1} 
\caption{
  % A schematic of the hierarchical model for combining
  % information across multiple worms in order to learn a shared
  % dynamical system for \celegans.  \textbf{a.}~All worms share a
  % canonical dynamics library. Here, we have three dynamics functions
  % (blue, red, gold), each of which is a simple linear mapping from
  % $x_t \in \reals^2$ to $x_{t+1} \in \reals^2$.  The recurrent model
  % assigns probabilities to the different dynamics functions based on
  % the current location in $x$-space.  \textbf{b.}~The worms also share
  % an emission matrix, $C$, which can be viewed as an embedding of
  % neurons in latent space.  \textbf{c.}~Each worm has its own dynamics
  % library, which allows for slight deviations from the canonical
  % dynamics.  \textbf{d.}~The model segments neural activity into a
  % sequence of discrete states~$z_{t} \in \{1, 2, 3\}$, which index
  % into the dynamics library.  \textbf{e.}~The discrete states specify
  % the dynamics that are used to propagate the continuous latent
  % states~$x_t$ forward in time. Here, the two dimensional latent
  % states are colored according to the instantaneous discrete
  % state. \textbf{f.}~The observed neural activity is a linear
  % projection of the underlying continuous latent states.  On any given
  % trial, we observe only a fraction of all the neurons.  The
  % hierarchical model combines these partial recordings to infer the
  % underlying states, the dynamics libraries, and the shared neural
  % embedding.
}
\label{fig:model}
\end{figure}

@@

% Robust model
While recurrent SLDS can approximate highly nonlinear systems, they
are still misspecified. \emph{Robust} methods~\citep{huber1981robust}
ameliorate this concern by lessening the penalty on observations that
do not quite match the dynamics of the current regime.  We achieve
this robustness by replacing the standard Gaussian noise model with a
multivariate-t distribution~\citep{lange1989robust}.  The heavy tails
of this distribution Fig.~\ref{fig:model}e compares a multivariate-t
distribution to a Gaussian with the same mean; though they have
the same covariance shape, the tails of the robust model fall off
more slowly than those of the Gaussian, allowing for outliers and a
degree of model misspecification.

% Hierarchical model
The third component of our framework is a \emph{hierarchical} model
for learning canonical dynamics from partial recordings collected from
many worms.  Since each recording contains a different subset of
labeled neurons, our first step is to align the data, treating the
activity of neurons that were not found in a given recording as
missing data that must be inferred.  We handle this by integrating
over possible activity of the missing neurons, analytically
marginalizing out this missing data.  The second challenge is
combining information from many recordings while also allowing for
worm-to-worm differences.  For example, the neural activitiy may have
larger variance in some recordings than in others due to different
levels of expression of the fluorescent protein.  We allow for this by
having a separate variance for each worm and neuron, as shown in
Fig.~\ref{fig:model}g.  More importantly, the dynamics of neural
activity may be slightly different from worm to worm.  We account for
this by introducing a canonical library of dynamical regimes
(Fig.~\ref{fig:model}f) that are shared by all worms as well as a
unique library for each individual worm (Fig.~\ref{fig:model}d);
however, we constrain the per-worm dynamics to be close to the
canonical dynamics.

% Inference
We fit the the canonical and per-worm dynamics libraries, the
parameters of the dynamics and observation noise models, and the
discrete and continuous latent states with approximate Bayesian
inference as described in the Supplementary Material. Briefly, we
separate the problem into two steps: learning the continuous latent
states and the mapping from continuous latent states to observations,
then learning the dynamics libraries that govern the continuous latent
states and the correspdonding discrete state sequences.  Both steps are
performed with a block Gibbs sampler, leveraging the structure of the
probabilistic model to perform efficient conditional updates of the
various parameters.

\subsection*{Emission model aids in identification and reveals clusters of neurons}

\begin{figure}[t!]
\centering%
\includegraphics[width=6in]{figures/figure2} 
\caption{}
\label{fig:emissions}
\end{figure}

The parameters of the learned model offer novel insights into the
structure of neural activity in the head ganglia that can only be
obtained by pooling information across many partial recordings.  With
this aggregated information, we can smooth noisy observations and
predict the activity of unobserved neurons. \citet{kato2015global}
used a smoothing procedure to estimate temporal principal components.
By contrast, we fit our model directly to the first-order differences
in calcium fluorescence; our only preprocessing step is to correct for
photo-bleaching. Fig~\ref{fig:emissions}a show our results for the first
of the five worms in this dataset. We see that the model smooths out
noise in the observed activity and, moreover, makes predictions about
the activity of neurons that could not be identified in this worm.

The predicted activity is based on the correlations between neurons
learned from many recordings.  These predictions are a useful aid for
identifying unlabeled neurons.  For example, we see that
\textsf{SMBDL} was not found in the first worm
(Fig.~\ref{fig:emissions}a). If the activity of one of the unlabeled
neurons closely matches the model's predictions for \textsf{SMBDL},
then this provides some evidence in favor of assigning it this
label. We tested this idea by witholding the labels of some
confidently-identified neurons and comparing their activity to the
model's predicted activity for all unlabeled neurons.  We measured the
similarity between the actual and predicted activity and found that
correct label, or ``candidate,'' tends to be much more similar than
other incorrect candidates, as summarized in
Fig.~\ref{fig:emissions}b.  Combined with other side-information, like
the location of the neuron, this provides a powerful aid for identifying
neurons.  The more neurons that we can identify, the better we will be
able to learn about the dynamics of neural activity.

Even though each individual recording contains only a subset of neurons,
the hierarchical model can still estimate the dimensionality of
neural activity in the entire head ganglia.  We reserve the last 20\% of
the data (3.6 min) for testing and estimate the dimensionality
of neural activity based on the marginal log likelihood the model assigns
to this held-out data. We find that the test likelihood peaks with a
ten-dimensional latent state (Fig.~\ref{fig:emissions}c).  Moreover, the likelihood is substantially
higher in the model that includes per-worm observation variances
compared to the model with shared variances for all worms. Fig.~\ref{fig:emissions}d
shows the inferred variance for each worm in the \citet{kato2015global}
dataset.  Neurons that were not identified are marked with a gray box.
This shows the extent of the missing data and the need for hierarchical
modeling.

The linear mapping from latent states to observations reveals clusters
of similarly tuned neurons.  The mapping maybe viewed as either an $N$
(neurons) by $D$ (latent dimensions) matrix or as a collection of~$N$
vectors in $D$-dimensional space, i.e. as an embedding of neurons.
These two views are shown in Fig.~\ref{fig:emissions}d. We note that
many vectors point in similar directions, which implies that there are
groups of correlated neurons.  We clustered the neurons based on the
cosine similarity between their embedding vecors and found five clusters
of neurons shown in Fig.~\ref{fig:emissions}f (order of neurons is the same
as in Fig.~\ref{fig:emissions}a). To make sense of these clusters, we then
looked at their spatial distribution along the head-to-tail axis of the
worm's body.  We found that clusters 2-4 correspond to activity closer to
the head and working backward toward the tail.  The last and largest cluster
contains many neurons in the nerve ring (location 0.15) \todo{check} as well
as some posterior neurons. \todo[inline]{discuss functional significance of clusters.}
